Explaining Black-Box Models through Counterfactuals

### 2024-01-07 examples/CounterfactualExplanations.jl/paper.tex

% Remove whitespace after paragraphs
\setlength{\parskip}{0pt}

\bibliographystyle{juliacon}
\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{biblatex}
\addbibresource{ref.bib}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Explaining Black-Box Models through Counterfactuals},
  pdfkeywords={Julia, Explainable Artificial
Intelligence, Counterfactual Explanations, Algorithmic Recourse},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Machine Learning models like Deep Neural Networks have become so complex
and opaque over recent years that they are generally considered
black-box systems. This lack of transparency exacerbates several other
problems typically associated with these models: they tend to be
unstable \autocite{goodfellow2014explaining}, encode existing biases
\autocite{buolamwini2018gender} and learn representations that are
surprising or even counter-intuitive from a human perspective
\autocite{buolamwini2018gender}. Nonetheless, they often form the basis
for data-driven decision-making systems in real-world applications.

As others have pointed out, this scenario gives rise to an undesirable
principal-agent problem involving a group of principals---i.e.~human
stakeholders---that fail to understand the behaviour of their
agent---i.e.~the black-box system \autocite{borch2022machine}. The group
of principals may include programmers, product managers and other
decision-makers who develop and operate the system as well as those
individuals ultimately subject to the decisions made by the system. In
practice, decisions made by black-box systems are typically left
unchallenged since the group of principals cannot scrutinize them:

\begin{quote}
``You cannot appeal to (algorithms). They do not listen. Nor do they
bend.'' \autocite{oneil2016weapons}
\end{quote}

@@

The field of XAI is still relatively young and made up of a variety of
subdomains, definitions, concepts and taxonomies. Covering all of these
is beyond the scope of this article, so we will focus only on high-level
concepts. The following literature surveys provide more detail: Arrieta
et al.~(2020) provide a broad overview of XAI
\autocite{arrieta2020explainable}; Fan et al.~(2020) focus on
explainability in the context of deep learning
\autocite{fan2020interpretability}; and finally, Karimi et al.~(2020)
\autocite{karimi2020survey} and Verma et al.~(2020)
\textcite{verma2020counterfactual} offer detailed reviews of the
literature on Counterfactual Explanations and Algorithmic Recourse (see
also \textcite{molnar2020interpretable} and
\textcite{varshney2022trustworthy}). Miller (2019) explicitly discusses
the concept of explainability from the perspective of a social scientist
\autocite{miller2019explanation}.

The first broad distinction we want to make here is between
\textbf{Interpretable} and \textbf{Explainable} AI. These terms are
often used interchangeably, but this can lead to confusion. We find the
distinction made in \textcite{rudin2019stop} useful: Interpretable AI
involves models that are inherently interpretable and transparent such
as general additive models (GAM), decision trees and rule-based models;
Explainable AI involves models that are not inherently interpretable but
require additional tools to be explainable to humans. Examples of the
latter include Ensembles, Support Vector Machines and Deep Neural
Networks. Some would argue that we best avoid the second category of
models altogether and instead focus solely on interpretable AI
\textcite{rudin2019stop}. While we agree that initial efforts should
always be geared towards interpretable models, avoiding black boxes
altogether would entail missed opportunities and anyway is probably not
very realistic at this point. For that reason, we expect the need for
XAI to persist in the medium term. Explainable AI can further be broadly
divided into \textbf{global} and \textbf{local} explainability: the
former is concerned with explaining the average behaviour of a model,
while the latter involves explanations for individual predictions
\autocite{molnar2020interpretable}. Tools for global explainability
include partial dependence plots (PDP), which involve the computation of
marginal effects through Monte Carlo, and global surrogates. A surrogate
model is an interpretable model that is trained to explain the
predictions of a black-box model.

@@

A recent study has shown that even experienced data scientists tend to
put too much trust in explanations produced by LIME and SHAP
\autocite{kaur2020interpreting}. Another recent work has shown that both
methods can be easily fooled: they depend on random input perturbations,
a property that can be abused by adverse agents to essentially whitewash
strongly biased black-box models \autocite{slack2020fooling}. In related
work, the same authors find that while gradient-based Counterfactual
Explanations can also be manipulated, there is a straightforward way to
protect against this in practice \autocite{slack2021counterfactual}. In
the context of quality assessment, it is also worth noting
that---contrary to surrogate explainers---CE always achieve full
fidelity by construction: counterfactuals are searched with respect to
the black-box classifier, not some proxy for it. That being said, CE
should also be used with care and research around them is still in its
early stages.

@@

Counterfactual search involves feature perturbations: we are interested
in understanding how we need to change individual attributes in order to
change the model output to a desired value or label
\autocite{molnar2020interpretable}. Typically the underlying methodology
is presented in the context of binary classification:
\(M: \mathcal{X} \mapsto \mathcal{Y}\) where
\(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
let \(t=1\) be the target class and let \(x\) denote the factual feature
vector of some individual sample outside of the target class, so
\(y=M(x)=0\). We follow this convention here, though it should be noted
that the ideas presented here also carry over to multi-class problems
and regression \autocite{molnar2020interpretable}.

The counterfactual search objective originally proposed by
\textcite{wachter2017counterfactual} is as follows

@@

Numerous extensions to this simple approach have been developed since CE
were first proposed in 2017 (see \textcite{verma2020counterfactual} and
\textcite{karimi2020survey} for surveys). The various approaches largely
differ in that they use different flavours of search objective defined
in Equation~\ref{eq-solution}. Different penalties are often used to
address many of the desirable properties of effective CE that have been
set out. These desiderata include: \textbf{proximity} --- the distance
between factual and counterfactual features should be small
\autocite{wachter2017counterfactual}; \textbf{actionability} --- the
proposed recourse should be actionable
\autocite{ustun2019actionable,poyiadzi2020face}; \textbf{plausibility}
--- the counterfactual explanation should be plausible to a human
\autocite{joshi2019realistic,schut2021generating}; \textbf{sparsity} ---
the counterfactual explanation should involve as few individual feature
changes as possible \autocite{schut2021generating}; \textbf{robustness}
--- the counterfactual explanation should be robust to domain and model
shifts \autocite{upadhyay2021robust}; \textbf{diversity} --- ideally
multiple diverse counterfactuals should be provided
\autocite{mothilal2020explaining}; and \textbf{causality} ---
counterfactuals should respect the structural causal model underlying
the data generating process
\autocite{karimi2020algorithmic,karimi2021algorithmic}.

@@

To the best of our knowledge, the package introduced here provides the
first implementation of Counterfactual Explanations in Julia and
therefore represents a novel contribution to the community. As for other
programming languages, we are only aware of one other unifying
framework: the Python library
\href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
\autocite{pawelczyk2021carla}.\footnote{While we were writing this
  paper, the \texttt{R} package \texttt{counterfactuals} was released
  \autocite{dandl2023counterfactuals}. The developers seem to also
  envision a unifying framework, but the project appears to still be in
  its early stages.} In addition to that, there exists open-source code
for some specific approaches to CE that have been proposed in recent
years. The approach-specific implementations that we have been able to
find are generally well-documented, but exclusively in Python. For
example, a PyTorch implementation of a greedy generator for Bayesian
models proposed in \textcite{schut2021generating} has been released. As
another example, the popular
\href{https://github.com/interpretml}{InterpretML} library includes an
implementation of a diverse counterfactual generator
\autocite{mothilal2020explaining}.

@@

The package currently offers native support for models built and trained
in \href{https://fluxml.ai/}{Flux} \autocite{innes2018flux} as well as a
small subset of models made available through
\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ}
\autocite{blaom2020mlj}. While in general it is assumed that users
resort to this package to explain their pre-trained models, we provide a
simple API call to train the following
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:

\begin{itemize}
\tightlist
\item
  Linear Classifier (Logistic Regression and Multinomial Logit)
\item
  Multi-Layer Perceptron (Deep Neural Network)
\item
  Deep Ensemble \textcite{lakshminarayanan2016simple}
\item
  Decision Tree, Random Forest, Gradient Boosted Trees
\end{itemize}

@@

A large and growing number of counterfactual
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/}{generators}
have already been implemented in our package (Table~\ref{tbl-gen}). At a
high level, we distinguish generators in terms of their compatible model
types, their default search space, and their composability. All
``gradient-based'' generators are compatible with differentiable models,
e.g.~\texttt{Flux} and \texttt{torch}, while ``tree-based'' generators
are only applicable to models that involve decision trees. Concerning
the search space, it is possible to search counterfactuals in a
lower-dimensional latent embedding of the feature space that implicitly
encodes the data-generating process (DGP). To learn the latent
embedding, existing work has typically relied on generative models or
existing causal knowledge
\autocite{joshi2019realistic,karimi2021algorithmic}. While this notion
is compatible with all of our gradient-based generators, only some
generators search a latent space by default. Finally, composability
implies that the given generator can be blended with any other
composable generator, which we discuss in Section~\ref{sec-gen-comp}.

Beyond these broad technical distinctions, generators largely differ in
terms of how they address the various desiderata mentioned above:
\emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
counterfactuals that are robust to endogenous model shifts
\autocite{altmeyer2023endogenous}; \emph{CLUE} searches plausible
counterfactuals in the latent embedding of a generative model by
explicitly minimising predictive entropy \autocite{antoran2020getting};
\emph{DiCE} is designed to generate multiple, maximally diverse
counterfactuals \autocite{mothilal2020explaining}; \emph{FeatureTweak}
leverages the internals of decision trees to search counterfactuals on a
feature-by-feature basis, finding the counterfactual that tweaks the
features in the least costly way \autocite{tolomei2017interpretable};
\emph{Gravitational} aims to generate plausible and robust
counterfactuals by minimising the distance to observed samples in the
target class \autocite{altmeyer2023endogenous}; \emph{Greedy} aims to
generate plausible counterfactuals by implicitly minimising predictive
uncertainty of Bayesian classifiers \autocite{schut2021generating};
\emph{GrowingSpheres} is model-agnostic, relying solely on identifying
nearest neighbours of counterfactuals in the target class by gradually
increasing the search radius and then moving counterfactuals in that
direction\autocite{laugel2017inversea}; \emph{PROBE} generates
probabilistically robust counterfactuals
\autocite{pawelczyk2022probabilistically}; \emph{REVISE} addresses the
need for plausibility by searching counterfactuals in the latent
embedding of a Variational Autoencoder (VAE)
\autocite{joshi2019realistic}; \emph{Wachter} is the baseline approach
that only penalises the distance to the original sample
\autocite{wachter2017counterfactual}.

\hypertarget{tbl-gen}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\caption{\label{tbl-gen}Overview of implemented counterfactual
generators.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ClaPROAR \autocite{altmeyer2023endogenous} & gradient based & feature &
yes \\
CLUE \autocite{antoran2020getting} & gradient based & latent & yes \\
DiCE \autocite{mothilal2020explaining} & gradient based & feature &
yes \\
FeatureTweak \autocite{tolomei2017interpretable} & tree based & feature
& no \\
Gravitational \autocite{altmeyer2023endogenous} & gradient based &
feature & yes \\
Greedy \autocite{schut2021generating} & gradient based & feature &
yes \\
GrowingSpheres \autocite{laugel2017inversea} & agnostic & feature &
no \\
PROBE \autocite{pawelczyk2022probabilistically} & gradient based &
feature & no \\
REVISE \autocite{joshi2019realistic} & gradient based & latent & yes \\
Wachter \autocite{wachter2017counterfactual} & gradient based & feature
& yes \\
\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  Adult Census \autocite{becker1996adult}
\item
  California Housing \autocite{pace1997sparse}
\item
  CIFAR10 \autocite{krizhevsky2009learning}
\item
  German Credit \autocite{hoffman1994german}
\item
  Give Me Some Credit \autocite{kaggle2011give}
\item
  MNIST \autocite{lecun1998mnist} and Fashion MNIST
  \autocite{xiao2017fashion}
\item
  UCI defaultCredit \autocite{yeh2009comparisons}
\end{itemize}

@@

where \(\mathbf{X}^\prime\) is an \(L\)-dimensional array of
counterfactuals,
\(f: \mathcal{Z}^{L \times K} \mapsto \mathcal{X}^{L \times D}\) is a
function that maps the \(L \times K\)-dimensional latent space
\(\mathcal{Z}\) to the \(L \times D\)-dimensional feature space
\(\mathcal{X}\) and \(\text{diversity}(\cdot)\) is the penalty proposed
by \textcite{mothilal2020explaining} that induces diverse sets of
counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the loss
function, \(M\) is the black-box model, \(t\) is the target class, and
\(\lambda\) is the strength of the penalty.

@@

The package also makes it easy to
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/evaluation/}{evaluate}
counterfactuals with respect to many of the desiderata mentioned above.
For example, users may want to infer how costly the provided recourse is
to individuals. To this end, we can measure the distance of the
counterfactual from its original value. The API call to compute the
distance metric defined in \textcite{wachter2017counterfactual}, for
instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
where \texttt{ce} can also be a vector of
\texttt{CounterfactualExplanation}s.

@@

To demonstrate how this can be done in practice, we will reiterate here
how native support for \href{https://fluxml.ai/}{\texttt{Flux.jl}}
\autocite{innes2018flux} deep learning models was enabled.\footnote{Flux
  models are now natively supported by our package and can be
  instantiated by calling \texttt{FluxModel()}.} Once again we use
synthetic data for an illustrative example. Code \ref{lst:nn} below
builds a simple model architecture that can be used for a multi-class
prediction task. Note how outputs from the final layer are not passed
through a softmax activation function, since the counterfactual loss is
evaluated with respect to logits as we discussed earlier. The model is
trained with dropout.

@@

\printbibliography[title=References]

>>>>

% Remove whitespace after paragraphs
\setlength{\parskip}{-1pt}
\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{juliacon.bst}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Explaining Black-Box Models through Counterfactuals},
  pdfkeywords={Julia, Explainable Artificial
Intelligence, Counterfactual Explanations, Algorithmic Recourse},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Machine Learning models like Deep Neural Networks have become so complex
and opaque over recent years that they are generally considered
black-box systems. This lack of transparency exacerbates several other
problems typically associated with these models: they tend to be
unstable \citep{goodfellow2014explaining}, encode existing biases
\citep{buolamwini2018gender} and learn representations that are
surprising or even counter-intuitive from a human perspective
\citep{buolamwini2018gender}. Nonetheless, they often form the basis for
data-driven decision-making systems in real-world applications.

As others have pointed out, this scenario gives rise to an undesirable
principal-agent problem involving a group of principals---i.e.~human
stakeholders---that fail to understand the behaviour of their
agent---i.e.~the black-box system \citep{borch2022machine}. The group of
principals may include programmers, product managers and other
decision-makers who develop and operate the system as well as those
individuals ultimately subject to the decisions made by the system. In
practice, decisions made by black-box systems are typically left
unchallenged since the group of principals cannot scrutinize them:

\begin{quote}
``You cannot appeal to (algorithms). They do not listen. Nor do they
bend.'' \citep{oneil2016weapons}
\end{quote}

@@

The field of XAI is still relatively young and made up of a variety of
subdomains, definitions, concepts and taxonomies. Covering all of these
is beyond the scope of this article, so we will focus only on high-level
concepts. The following literature surveys provide more detail: Arrieta
et al.~(2020) provide a broad overview of XAI
\citep{arrieta2020explainable}; Fan et al.~(2020) focus on
explainability in the context of deep learning
\citep{fan2020interpretability}; and finally, Karimi et al.~(2020)
\citep{karimi2020survey} and Verma et al.~(2020)
\citet{verma2020counterfactual} offer detailed reviews of the literature
on Counterfactual Explanations and Algorithmic Recourse (see also
\citet{molnar2020interpretable} and \citet{varshney2022trustworthy}).
Miller (2019) explicitly discusses the concept of explainability from
the perspective of a social scientist \citep{miller2019explanation}.

The first broad distinction we want to make here is between
\textbf{Interpretable} and \textbf{Explainable} AI. These terms are
often used interchangeably, but this can lead to confusion. We find the
distinction made in \citet{rudin2019stop} useful: Interpretable AI
involves models that are inherently interpretable and transparent such
as general additive models (GAM), decision trees and rule-based models;
Explainable AI involves models that are not inherently interpretable but
require additional tools to be explainable to humans. Examples of the
latter include Ensembles, Support Vector Machines and Deep Neural
Networks. Some would argue that we best avoid the second category of
models altogether and instead focus solely on interpretable AI
\citet{rudin2019stop}. While we agree that initial efforts should always
be geared towards interpretable models, avoiding black boxes altogether
would entail missed opportunities and anyway is probably not very
realistic at this point. For that reason, we expect the need for XAI to
persist in the medium term. Explainable AI can further be broadly
divided into \textbf{global} and \textbf{local} explainability: the
former is concerned with explaining the average behaviour of a model,
while the latter involves explanations for individual predictions
\citep{molnar2020interpretable}. Tools for global explainability include
partial dependence plots (PDP), which involve the computation of
marginal effects through Monte Carlo, and global surrogates. A surrogate
model is an interpretable model that is trained to explain the
predictions of a black-box model.

@@

A recent study has shown that even experienced data scientists tend to
put too much trust in explanations produced by LIME and SHAP
\citep{kaur2020interpreting}. Another recent work has shown that both
methods can be easily fooled: they depend on random input perturbations,
a property that can be abused by adverse agents to essentially whitewash
strongly biased black-box models \citep{slack2020fooling}. In related
work, the same authors find that while gradient-based Counterfactual
Explanations can also be manipulated, there is a straightforward way to
protect against this in practice \citep{slack2021counterfactual}. In the
context of quality assessment, it is also worth noting that---contrary
to surrogate explainers---CE always achieve full fidelity by
construction: counterfactuals are searched with respect to the black-box
classifier, not some proxy for it. That being said, CE should also be
used with care and research around them is still in its early stages.

@@

Counterfactual search involves feature perturbations: we are interested
in understanding how we need to change individual attributes in order to
change the model output to a desired value or label
\citep{molnar2020interpretable}. Typically the underlying methodology is
presented in the context of binary classification:
\(M: \mathcal{X} \mapsto \mathcal{Y}\) where
\(\mathcal{X}\subset\mathbb{R}^D\) and \(\mathcal{Y}=\{0,1\}\). Further,
let \(t=1\) be the target class and let \(x\) denote the factual feature
vector of some individual sample outside of the target class, so
\(y=M(x)=0\). We follow this convention here, though it should be noted
that the ideas presented here also carry over to multi-class problems
and regression \citep{molnar2020interpretable}.

The counterfactual search objective originally proposed by
\citet{wachter2017counterfactual} is as follows

@@

Numerous extensions to this simple approach have been developed since CE
were first proposed in 2017 (see \citet{verma2020counterfactual} and
\citet{karimi2020survey} for surveys). The various approaches largely
differ in that they use different flavours of search objective defined
in Equation~\ref{eq-solution}. Different penalties are often used to
address many of the desirable properties of effective CE that have been
set out. These desiderata include: \textbf{proximity} --- the distance
between factual and counterfactual features should be small
\citep{wachter2017counterfactual}; \textbf{actionability} --- the
proposed recourse should be actionable
\citep{ustun2019actionable, poyiadzi2020face}; \textbf{plausibility} ---
the counterfactual explanation should be plausible to a human
\citep{joshi2019realistic, schut2021generating}; \textbf{sparsity} ---
the counterfactual explanation should involve as few individual feature
changes as possible \citep{schut2021generating}; \textbf{robustness} ---
the counterfactual explanation should be robust to domain and model
shifts \citep{upadhyay2021robust}; \textbf{diversity} --- ideally
multiple diverse counterfactuals should be provided
\citep{mothilal2020explaining}; and \textbf{causality} ---
counterfactuals should respect the structural causal model underlying
the data generating process
\citep{karimi2020algorithmic, karimi2021algorithmic}.

@@

To the best of our knowledge, the package introduced here provides the
first implementation of Counterfactual Explanations in Julia and
therefore represents a novel contribution to the community. As for other
programming languages, we are only aware of one other unifying
framework: the Python library
\href{https://carla-counterfactual-and-recourse-library.readthedocs.io/en/latest/?badge=latest}{CARLA}
\citep{pawelczyk2021carla}.\footnote{While we were writing this paper,
  the \texttt{R} package \texttt{counterfactuals} was released
  \citep{dandl2023counterfactuals}. The developers seem to also envision
  a unifying framework, but the project appears to still be in its early
  stages.} In addition to that, there exists open-source code for some
specific approaches to CE that have been proposed in recent years. The
approach-specific implementations that we have been able to find are
generally well-documented, but exclusively in Python. For example, a
PyTorch implementation of a greedy generator for Bayesian models
proposed in \citet{schut2021generating} has been released. As another
example, the popular \href{https://github.com/interpretml}{InterpretML}
library includes an implementation of a diverse counterfactual generator
\citep{mothilal2020explaining}.

@@

The package currently offers native support for models built and trained
in \href{https://fluxml.ai/}{Flux} \citep{innes2018flux} as well as a
small subset of models made available through
\href{https://alan-turing-institute.github.io/MLJ.jl/dev/}{MLJ}
\citep{blaom2020mlj}. While in general it is assumed that users resort
to this package to explain their pre-trained models, we provide a simple
API call to train the following
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/model_catalogue/}{models}:

\begin{itemize}
\tightlist
\item
  Linear Classifier (Logistic Regression and Multinomial Logit)
\item
  Multi-Layer Perceptron (Deep Neural Network)
\item
  Deep Ensemble \citet{lakshminarayanan2016simple}
\item
  Decision Tree, Random Forest, Gradient Boosted Trees
\end{itemize}

@@

A large and growing number of counterfactual
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/explanation/generators/overview/}{generators}
have already been implemented in our package (Table~\ref{tbl-gen}). At a
high level, we distinguish generators in terms of their compatible model
types, their default search space, and their composability. All
``gradient-based'' generators are compatible with differentiable models,
e.g.~\texttt{Flux} and \texttt{torch}, while ``tree-based'' generators
are only applicable to models that involve decision trees. Concerning
the search space, it is possible to search counterfactuals in a
lower-dimensional latent embedding of the feature space that implicitly
encodes the data-generating process (DGP). To learn the latent
embedding, existing work has typically relied on generative models or
existing causal knowledge
\citep{joshi2019realistic, karimi2021algorithmic}. While this notion is
compatible with all of our gradient-based generators, only some
generators search a latent space by default. Finally, composability
implies that the given generator can be blended with any other
composable generator, which we discuss in Section~\ref{sec-gen-comp}.

Beyond these broad technical distinctions, generators largely differ in
terms of how they address the various desiderata mentioned above:
\emph{ClapROAR} aims to preserve the classifier, i.e.~to generate
counterfactuals that are robust to endogenous model shifts
\citep{altmeyer2023endogenous}; \emph{CLUE} searches plausible
counterfactuals in the latent embedding of a generative model by
explicitly minimising predictive entropy \citep{antoran2020getting};
\emph{DiCE} is designed to generate multiple, maximally diverse
counterfactuals \citep{mothilal2020explaining}; \emph{FeatureTweak}
leverages the internals of decision trees to search counterfactuals on a
feature-by-feature basis, finding the counterfactual that tweaks the
features in the least costly way \citep{tolomei2017interpretable};
\emph{Gravitational} aims to generate plausible and robust
counterfactuals by minimising the distance to observed samples in the
target class \citep{altmeyer2023endogenous}; \emph{Greedy} aims to
generate plausible counterfactuals by implicitly minimising predictive
uncertainty of Bayesian classifiers \citep{schut2021generating};
\emph{GrowingSpheres} is model-agnostic, relying solely on identifying
nearest neighbours of counterfactuals in the target class by gradually
increasing the search radius and then moving counterfactuals in that
direction\citep{laugel2017inversea}; \emph{PROBE} generates
probabilistically robust counterfactuals
\citep{pawelczyk2022probabilistically}; \emph{REVISE} addresses the need
for plausibility by searching counterfactuals in the latent embedding of
a Variational Autoencoder (VAE) \citep{joshi2019realistic};
\emph{Wachter} is the baseline approach that only penalises the distance
to the original sample \citep{wachter2017counterfactual}.

\hypertarget{tbl-gen}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2500}}@{}}
\caption{\label{tbl-gen}Overview of implemented counterfactual
generators.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Generator
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Model Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Search Space
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Composable
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
ClaPROAR \citep{altmeyer2023endogenous} & gradient based & feature &
yes \\
CLUE \citep{antoran2020getting} & gradient based & latent & yes \\
DiCE \citep{mothilal2020explaining} & gradient based & feature & yes \\
FeatureTweak \citep{tolomei2017interpretable} & tree based & feature &
no \\
Gravitational \citep{altmeyer2023endogenous} & gradient based & feature
& yes \\
Greedy \citep{schut2021generating} & gradient based & feature & yes \\
GrowingSpheres \citep{laugel2017inversea} & agnostic & feature & no \\
PROBE \citep{pawelczyk2022probabilistically} & gradient based & feature
& no \\
REVISE \citep{joshi2019realistic} & gradient based & latent & yes \\
Wachter \citep{wachter2017counterfactual} & gradient based & feature &
yes \\
\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  Adult Census \citep{becker1996adult}
\item
  California Housing \citep{pace1997sparse}
\item
  CIFAR10 \citep{krizhevsky2009learning}
\item
  German Credit \citep{hoffman1994german}
\item
  Give Me Some Credit \citep{kaggle2011give}
\item
  MNIST \citep{lecun1998mnist} and Fashion MNIST \citep{xiao2017fashion}
\item
  UCI defaultCredit \citep{yeh2009comparisons}
\end{itemize}

@@

where \(\mathbf{X}^\prime\) is an \(L\)-dimensional array of
counterfactuals,
\(f: \mathcal{Z}^{L \times K} \mapsto \mathcal{X}^{L \times D}\) is a
function that maps the \(L \times K\)-dimensional latent space
\(\mathcal{Z}\) to the \(L \times D\)-dimensional feature space
\(\mathcal{X}\) and \(\text{diversity}(\cdot)\) is the penalty proposed
by \citet{mothilal2020explaining} that induces diverse sets of
counterfactuals. As in Equation~\ref{eq-solution}, \(\ell\) is the loss
function, \(M\) is the black-box model, \(t\) is the target class, and
\(\lambda\) is the strength of the penalty.

@@

The package also makes it easy to
\href{https://juliatrustworthyai.github.io/CounterfactualExplanations.jl/v0.1/tutorials/evaluation/}{evaluate}
counterfactuals with respect to many of the desiderata mentioned above.
For example, users may want to infer how costly the provided recourse is
to individuals. To this end, we can measure the distance of the
counterfactual from its original value. The API call to compute the
distance metric defined in \citet{wachter2017counterfactual}, for
instance, is as simple as \texttt{evaluate(ce;\ measure=distance\_mad)},
where \texttt{ce} can also be a vector of
\texttt{CounterfactualExplanation}s.

@@

To demonstrate how this can be done in practice, we will reiterate here
how native support for \href{https://fluxml.ai/}{\texttt{Flux.jl}}
\citep{innes2018flux} deep learning models was enabled.\footnote{Flux
  models are now natively supported by our package and can be
  instantiated by calling \texttt{FluxModel()}.} Once again we use
synthetic data for an illustrative example. Code \ref{lst:nn} below
builds a simple model architecture that can be used for a multi-class
prediction task. Note how outputs from the final layer are not passed
through a softmax activation function, since the counterfactual loss is
evaluated with respect to logits as we discussed earlier. The model is
trained with dropout.

@@

\renewcommand\refname{References}
  \bibliography{ref.bib}
